---
title: "Demo-DH. Vergelijking meetgegevens RIVM (AVK, BL, RS)"
author: "Gerrit Versteeg"
date: "January 31th, 2017"
output: 
        html_document:
                keep_md: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
In dit rapport onderzoeken we de verschillen in de metingen voor drie RIVM-meetstations locaties in Den Haag (Amsterdamse Veerkade, Bleriotlaat en Rebecquestraat).

De analyse leidt to de volgende conclusies:


## Data Processing
### Loading the data
De data voor dit onderzoek is opgevraagd bij: https://www.luchtmeetnet.nl/download
door selectie van telkens één van de vier RIVM-meetstations in Den Haag,
gevolgd door de selectie van de breedste periode van aanwezige gevalideerde meetdata (groen gemarkeerde datums) en de invoer van een email-adres voor het resulterende csv-bestand (extract.csv). De csv-bestanden zijn vanuit de eMails opgeslagen in de werkdirectory onder de subdirectory 'data'.

Dit proces is uitgevoerd voor alle (4) meetstations in Den Haag:
Export_AVK.csv (312 kB) - Amsterdamse Veerkade
Export_BL.csv  (353 kB) - Bleriotlaan
Export_RS.csv  (468 kB) - Rebecquestraat
Vaillantlaan, kent geen datums met gevalideerde meetdata en levert derhalve een leeg bestand op.

De drie files zijn in Nederlands csv-formaat (sep=; en dec=,) daarom gebruiken we read.csv2. De files worden in tibbles geplaatst en daarna samengevoegd tot één tibble (DF_RIVM) gebruikmakend van dplyr voor snellere en eenvoudigere datamanipulatie.

```{r loading,ECHO = TRUE}
library("plyr", warn.conflicts=FALSE)           ## load plyr silently
library("dplyr", warn.conflicts=FALSE)          ## load dplyr silently
library("ggplot2", warn.conflicts=FALSE)        ## load ggplot2 silently

dateDownloaded <- date()                        ## register date of download
DF_AVK <- tbl_df(read.csv2("./data/Export_AVK.csv"))
DF_BL <- tbl_df(read.csv2("./data/Export_BL.csv"))
DF_RS <- tbl_df(read.csv2("./data/Export_RS.csv"))
DF_RIVM <- bind_rows(DF_AVK, DF_BL, DF_RS)      ## concatenate station data
DF_RIVM                                         ## take a peek
```

Afgeleid uit de door LMN meegestuurde data-sheet bevat het bestand gevalideerde uurgemiddelde meetwaarden in csv-format (punt-komma gescheiden):

* Tijdstip; eindtijdstip van metingen van metingen waarmee uurgemiddelde is bepaald
* Locatie; naam van meetpunt
* Component; naam van component
* Waarde; uurgemiddelde meetwaarde (decimaalteken is een komma)
* LKI (Luchtkwaliteitsindex); meetwaarde is omgezet tot een getal van 1 (weinig luchtverontreiniging) tot 11 (veel luchtverontreiniging), voor meer informatie zie https://www.luchtmeetnet.nl/uitleg#

Kijkend naar de 'structure' van de drie meetbestanden:

* AVK kent 1730 meetmomenten voor NO, NO2 en PM10
* BL kent 1751 meetmomenten voor NO, NO2, O3 en PM10
* RS kent 1806 meetmomenten voor NO, NO2, O3, PM10 en PM2.5

In dit voorbeeld kiezen we ervoor om alleen de componenten en meetmomenten te nemen die in alle drie meetstations voorkomen.

### Preprocessing the data
Next step is to pre-process the data for analysis.  

**tijdstip**
Deze variabele geeft ons het meetmoment. Hierbij geldt dat:
2016-01-01 01:00:00+01:00 het gemeten resultaat is van de periode tussen 2016-01-01 00:00:00 en 2016-01-01 01:00:00 in wintertijd in Nederland.

Laten we eerst kijken naar de hoeveelheid unieke meetmomenten per meetstation:
```{r levelcheck_tijdstip, ECHO=TRUE}
length(levels(DF_AVK$tijdstip))
length(levels(DF_BL$tijdstip))
length(levels(DF_RS$tijdstip))
```

Het aantal meetmomenten verschilt per meetstation en daar moeten we rekening mee houden want voor onderlinge vergelijking moeten we alleen de meetmomenten hebben met registraties voor elke van de drie stations.


Vanuit de variabele tijdstip gaan we in verband met plotting in de tijd twee nieuwe variabelen maken, namelijk: datetime (POSIXct-formaat), date (Date-formaat) en time (Time-formaat). Hiermee kunnen individueel meetmomenten herkennen en ook gemiddelden per dag met elkaar vergelijken en de meetwaarden over de verschillende momenten op een dag.

**locatie**
Het betrokken meetstation gaan we afkorten tot een handzamer formaat: AVK, BL, RS en in factor formaat zetten.

**component**
De variabele "component" verwijst naar de gemeten indicator (NO, NO2, O3, PM10 en PM2.5). Dit wordt een factor-formaat. Laten we even kijken naar welke indicatoren door welk meetstation worden gemeten.

```{r EXPcheck, ECHO=TRUE}
levels(DF_AVK$component)
levels(DF_BL$component)
levels(DF_RS$component)
```

Alle drie meten NO, NO2 en PM10. Bleriotlaan meet ook nog O3, terwijl de Rebecquestraat ook O3 en PM2.5 meet. Ook hier moeten we rekening meehouden.

**waarde**
Is de werkelijke meetwaarde als een uurgemiddelde over het meetmoment in ug/m3. Formaat houden we numeric.


**LKI**
Is een categorisering van de meetwaarden in waarden van 1 (goed) tot 11 (slecht). Deze variabele gebruiken we voorlopig niet.


```{r Preprocessing,ECHO=TRUE}

gmm <- intersect(levels(DF_AVK$tijdstip),           ## vul gmm met de gemeen-
              intersect(levels(DF_BL$tijdstip),     ## schappelijke meetmomenten
                levels(DF_RS$tijdstip)))            ## in alle drie meetstations

gcp <- intersect(levels(DF_AVK$component),          ## vul gcp met de gemeen-
              intersect(levels(DF_BL$component),    ## schappelijke indicatoren
                levels(DF_RS$component)))           ## in alle drie meetstations

DF_prep <-                                          ## Create DF_prep
        DF_RIVM %>%                                 ## use DF-raw to
        filter(tijdstip %in% gmm) %>%               ## alleen gemeensch datums
        filter(component %in% gcp) %>%              ## alleen gemeensch indicatoren
        mutate(locatie = as.factor(locatie)) %>%    ## set formaat factor
        mutate(component = as.factor(component)) %>%  ## set formaat factor
        mutate(datetime = as.POSIXct(strptime(tijdstip,
                "%Y-%m-%d %H:%M:%S"))) %>%          ## add datetime column
        mutate(date = as.Date(strptime(tijdstip, 
                "%Y-%m-%d"))) %>%                   ## add date column
                print()                                     ## and let's have a look
```

Even controleren of we geen missing values hebben door de combinatie van drie meetstations:

```{r checkNA, ECHO = TRUE}
sum(is.na(DF_prep))                 ## to check for missing values in dataset
```

No missing values are present, so we can start doing the analysis.

## Results
### Plotting the indicators in time
We gaan de drie indicatoren (NO, NO2 en PM10) in drie plots met verschillende kleuren voor de indicatoren per maand uitzetten in een plot, om de verschillen te bekijken. 

```{r prep4plot1, ECHO=TRUE}
DF_month <-                                              ## create DF_month
        DF_prep %>%                                      ## using DF-prep to
        group_by(date, component, locatie) %>%           ## group by date,comp,loc
        summarize(avg = mean(waarde, na.rm = TRUE)) %>%  ## calc. avg waarde
        arrange(date)                                    ## sort ascending by date
```

Now we will create a figure that shows two tranposed barplots to visualize the results for the first 10 types of events that cause the highest harm to the population; one for the average harm and another for the total harm. 

```{r month-plot, ECHO=TRUE}
g <- ggplot(DF_month, 
        aes(x=date, y=avg))                             ## setup graphic object

g   + geom_point(aes(color=component)) +                ## plot the points
      geom_smooth(aes(color=component), method="lm") +                        ## plot a trendline
      facet_grid(locatie~.) +                   ## plot facets per comp/loc
      xlab("Tijd") +                                    ## lable X-axis
      ylab("Emissie") +                                 ## lable Y-axis
      ggtitle("Emissies in Den Haag (2015-2017)")       ## title plot
```

Looking at the *total* harm caused by severe weather events (bottom barplot), we see 'TORNADO' having the highest harmfulness by far, with almost 100.000 people harmed (fatalities + injuries). 

> **Average harmfulness per type of event**


| EVTYPE | People harmed on average |
|----|:----|
| `r DF_harmavg[1,1]` | `r as.integer(DF_harmavg[1,2])` |
| `r DF_harmavg[2,1]` | `r as.integer(DF_harmavg[2,2])` |
| `r DF_harmavg[3,1]` | `r as.integer(DF_harmavg[3,2])` |
| `r DF_harmavg[4,1]` | `r as.integer(DF_harmavg[4,2])` |
| `r DF_harmavg[5,1]` | `r as.integer(DF_harmavg[5,2])` |
| `r DF_harmavg[6,1]` | `r as.integer(DF_harmavg[6,2])` |
| `r DF_harmavg[7,1]` | `r as.integer(DF_harmavg[7,2])` |
| `r DF_harmavg[8,1]` | `r as.integer(DF_harmavg[8,2])` |
| `r DF_harmavg[9,1]` | `r as.integer(DF_harmavg[9,2])` |
| `r DF_harmavg[10,1]` | `r as.integer(DF_harmavg[10,2])` |

Harmfulness per event type *on average* however, shows 'Heat Waves' (70 people harmed) as causing the most fatalities and injuries, followed by 'TROPICAL STORM GORDON'(51 people harmed). 

The plot shows us that there is a problem with the event types in the dataset, for instance: tropical storm gordon is a specific storm and not the event type 'tropical storm'. Types and occurrences seem to have been mixed up. Another issue is the 'Heat Wave' versus "HEAT WAVE', where event types have been seperated due to their spelling and typecase.
Note that we will **not** clean this up, because the assignment clearly states that we need to use the EVTYPE variable as a given.


### Economic Consequences
In this section we analyze which types of events (as indicated in the EVTYPE variable) have the most severe economic consequences, across the United States.
Events can have economic consequences (damage) to property as well as crops. Because no definiton of 'economic consequence' is prescribed and no reference is made to separate damage to crops from damage to property, we will simply consider the sum of property damage and crop damage to be the total 'economic consequence' of the event type. The total damage will be put in a new variable 'DAMAGE'.

We will continu to use the average of the damage next to the total damage for each event type, as explained under 'unbalanced registration' in the section Data Analysis, Harmful Events. 
Using DAMAGE we calculate the total damage per event type (TOT) and the mean per type of event (AVG), to determine which events have the highest damage on average. Event types with an average or total damage of 0 will be dropped.

```{r deriveDamage, ECHO=TRUE}
DF_damavg <-                                          ## create DF_damavg
        DF_prep %>%                                   ## use DF-prep to
        mutate(DAMAGE = PROPDMGTOT + CROPDMGTOT) %>%  ## add the column DAMAGE
        group_by(EVTYPE) %>%                          ## group by EVTYPE
        summarize(AVG = mean(DAMAGE)) %>%             ## calc. avg damage
        mutate(AVG = AVG / 1000000000) %>%            ## in billions of dollars
        arrange(desc(AVG)) %>%                        ## sort descending by AVG
        filter(AVG > 0)

DF_damsum <-                                          ## create DF_damsum
        DF_prep %>%                                   ## use DF-prep to
        mutate(DAMAGE = PROPDMGTOT + CROPDMGTOT) %>%  ## add the column DAMAGE
        group_by(EVTYPE) %>%                          ## group by EVTYPE
        summarize(TOT = sum(DAMAGE)) %>%              ## calc. total damage
        mutate(TOT = TOT / 1000000000) %>%            ## in billions of dollars
        arrange(desc(TOT)) %>%                        ## sort descending by total
        filter(TOT > 0)
```

Now we will generate a figure with two tranposed barplots to visualize the economic damage caused by the first highest-ranking 10 events (crops + properties). The top barplot shows the avarage damage per event type, while the bottom barplot depicts the average damage for each type of event. 

```{r damage-plot, ECHO=TRUE}
d <- ggplot(DF_damsum[1:10,], 
        aes(x=reorder(EVTYPE, TOT), y = TOT))            ## setup graphic object
d+geom_bar(stat = "identity") +                          ## plot the bars
        coord_flip() +                                   ## flip plot
        xlab("Type of event") +                          ## lable X-axis
        ylab("Total economic damage (billions USD)") +   ## lable Y-axis
        ggtitle("Damage causes by Weather Events (1950-2011)") ## title plot
```

Looking at the total damage on crops and property caused by severe weather events, 'FLOOD' ranks highest with 150 billion USD, followed by 'HURRICANE/TYPHOON' (72 billion USD) and 'TORNADO' (53 billion USD).

However, looking at the table below, event type 'TORNADOES, TSTM WIND, HAIL' causes the most damage on average: 1.6 billion USD. Followed by 'HEAVY RAIN/SEVERE WEATHER' with average damages of 1.25 billion USD.


> **Average damage per type of event**

| EVTYPE | Damage on average (billions of dollars) |
|----|:----|
| `r DF_damavg[1,1]` | `r round(DF_damavg[1,2], digits=3)` |
| `r DF_damavg[2,1]` | `r round(DF_damavg[2,2], digits=3)` |
| `r DF_damavg[3,1]` | `r round(DF_damavg[3,2], digits=3)` |
| `r DF_damavg[4,1]` | `r round(DF_damavg[4,2], digits=3)` |
| `r DF_damavg[5,1]` | `r round(DF_damavg[5,2], digits=3)` |
| `r DF_damavg[6,1]` | `r round(DF_damavg[6,2], digits=3)` |
| `r DF_damavg[7,1]` | `r round(DF_damavg[7,2], digits=3)` |
| `r DF_damavg[8,1]` | `r round(DF_damavg[8,2], digits=3)` |
| `r DF_damavg[9,1]` | `r round(DF_damavg[9,2], digits=3)` |
| `r DF_damavg[10,1]` | `r round(DF_damavg[10,2], digits=3)` |


Again we see the awkward use of types versus occurrences. Hurricanes are listed as type 'HURRICANE', as well as occurrence 'HURRICANE OPAL' and even mixed types like 'HURRICANE OPAL/HIGH WINDS'. Nevertheless, as stated before, we kept on using the EVTYPE **as such** and did not transform EVTYPES into nicely belanced actual weather types.


## Warning regarding the value of the results
The variable EVTYPE in the given dataset (1995-2011) is not defined and categorized correctly. Types and occurrences are mixed in together, while typecasing leads to unnecessary variation in event types. In essence this means that the conclusions only refer to 'EVTYPE', i.e. NOT to actual types of weather events.